---
title: "BIOL343 Coding Challenge 3"
author: "ANSWER KEY — for TAs, not for OnQ"
date: "2026-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BIOL343 Coding Challenge 3

## Group Members

```{r group_members}
# Name - Queen's Student Number

# Key drafted 2026-09-21 (hospital scenario; the lakes scenario is identical code with different column names)

```

**Scenario chosen:** hospital.

# DO THIS FIRST {-}

Copy and paste this into your R Console to start positpal, but replace `FILENAME.Rmd` with the name you are going to use for the .R or .Rmd file you are going to upload (rename the template first, e.g. `Group_#_CC3.Rmd`, and use that exact name). **Do not include it as a code chunk.**

```
install.packages("remotes")
remotes::install_url("https://quantitative.bio/downloads/positpal.tar.gz", upgrade = "never")
positpal::start("FILENAME.Rmd")
```

Before you run it: open the project folder for this challenge in Positron (*File → Open Folder…*) so that it is your working directory, save this file in that folder under the name you will upload, and check that autosave is on (*File → Auto Save*). The first time you use positpal on a computer, `start()` will ask you to run `positpal::accept_eula()` once; do that, then run the `start()` line again. After a few minutes of work, check in the Explorer pane that the `.capture/snaps` folder inside your project is filling up with files. If it is not, or if `start()` refuses to record, run `positpal::doctor("FILENAME.Rmd")` and bring the output to tutorial or office hours. Run the `start()` line again at the start of every session you spend on this file.

# Part 1 — Import and inspect three related files

## 1. Load packages

```{r load-packages, message=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(readr)
library(ggplot2)
```

`library(tidyverse)` is equally fine. Grading: any set of `library()` calls that makes the rest run.

## 2. Import the three files

```{r import-data}
data1 <- read_csv("data/Dataset1.csv", show_col_types = FALSE)
data2 <- read_csv("data/Dataset2.csv", show_col_types = FALSE)
data3 <- read_delim("data/Dataset3.csv", delim = ";", show_col_types = FALSE)
```

Accept `read.csv("data/Dataset3.csv", sep = ";")` too; the difference shows up in Section 6 (`Date` stays character with `read.csv()`). `read_csv2()` is **wrong** here: it expects a comma as the decimal mark, and this file uses a period, so the PC scores would be mangled.

## 3. Inspect the data

```{r inspect-data}
str(data1)
str(data2)
str(data3)
```

**Question 1:** Which columns did R import as a type you did not expect, and what in the raw files caused it?

**Answer:** `PC1` and `PC2` are character (`chr`) in `data2` and `data3`, not numeric. `data2` has the word `Missing` in some cells and `data3` has `N/A`; one non-numeric cell is enough to make R read the whole column as text. (If `read.csv()` was used, `Date` is also character.) `Conc` in `data1` is numeric even though some values are written in scientific notation.

## 4. Fix the missing values

```{r fix-missing}
data2 <- data2 %>%
  mutate(PC1 = as.numeric(PC1), PC2 = as.numeric(PC2))
data3 <- data3 %>%
  mutate(PC1 = as.numeric(PC1), PC2 = as.numeric(PC2))

sum(is.na(data2$PC1))
sum(is.na(data2$PC2))
sum(is.na(data3$PC1))
sum(is.na(data3$PC2))
```

Expect 8, 4, 7 and 4. Two coercion warnings are expected. Also accept re-importing with `na = c("", "NA", "Missing", "N/A")` (readr) or `na.strings = c("NA", "Missing", "N/A")` (`read.csv()`), which gives numeric columns directly.

## 5. Rename the measurement columns

```{r rename-columns}
data2 <- data2 %>% rename(micro_PC1 = PC1, micro_PC2 = PC2)
data3 <- data3 %>% rename(urine_PC1 = PC1, urine_PC2 = PC2)
names(data2)
names(data3)
```

**Question 2:** What would have happened to these column names if you had joined without renaming?

**Answer:** The join keeps both and disambiguates them with suffixes: `PC1.x`, `PC2.x` (from the first table) and `PC1.y`, `PC2.y` (from the second). The code still runs, but nothing in the names says which file is which, which is how errors of interpretation start.

## 6. Make sure `Date` is a date

```{r fix-dates}
class(data1$Date)
class(data2$Date)
class(data3$Date)
data1 <- data1 %>% mutate(Date = ymd(Date))
data2 <- data2 %>% mutate(Date = ymd(Date))
data3 <- data3 %>% mutate(Date = ymd(Date))
class(data3$Date)
```

With `readr` the dates are already `Date`, and `ymd()` on a `Date` is harmless. With `read.csv()` they are character and the join in Section 7 fails with "Can't join ... due to incompatible types (character / date)" until this step is done. Grading: the check plus a conversion where needed.

# Part 2 — Join the files

## 7. Full join by ID and Date

```{r full-join}
all_data <- full_join(data1, data2, by = c("ID", "Date")) %>%
  full_join(data3, by = c("ID", "Date"))
dim(all_data)
```

800 rows, 7 columns (`ID`, `Date`, `Conc`, four PC columns, and no duplicates). Joining by `ID` alone is a common error: each ID appears in two years, so the result has 1,600 rows and pairs 2021 samples with 2022 samples. Accept nested calls instead of the pipe.

## 8. Compare the join verbs

```{r compare-joins}
nrow(inner_join(data1, data2, by = c("ID", "Date")))
nrow(left_join(data1, data2, by = c("ID", "Date")))
nrow(right_join(data1, data2, by = c("ID", "Date")))
nrow(full_join(data1, data2, by = c("ID", "Date")))
```

**Question 3:** Explain which rows each verb keeps.

**Answer:** `data1` has 794 samples and `data2` has 788; 782 are in both. `inner_join()` keeps only the 782 matches. `left_join()` keeps every row of `data1` (794) and fills the missing microbiome scores with `NA`. `right_join()` keeps every row of `data2` (788) and fills `Conc` with `NA` where `data1` has no sample. `full_join()` keeps everything (800), with `NA` on whichever side is missing. The samples that were measured for one thing but not the other survive only in the left/right/full results, so which verb you choose decides whether those patients (lakes) are silently dropped.

# Part 3 — Dates, summaries and ordering

## 9. Add year and month

```{r add-year}
all_data <- all_data %>%
  mutate(Year = year(Date), Month = month(Date, label = TRUE))
head(all_data)
```

`month(Date)` (a number) is fine too.

## 10. Summarise concentration by year

```{r year-summary}
year_summary <- all_data %>%
  group_by(Year) %>%
  summarise(mean_conc = mean(Conc, na.rm = TRUE),
            sd_conc = sd(Conc, na.rm = TRUE))
year_summary
```

**Question 4:** Does concentration differ between the two years?

**Answer:** Yes, strongly. Mean concentration is about 10.2 in 2021 (SD 6.1) and about 1.0 in 2022 (SD 1.5), a roughly ten-fold drop, far larger than the spread within either year. Any answer that reads the two rows of the table correctly gets credit; a formal test comes later in the course.

## 11. The five highest concentrations

```{r top-five}
all_data %>%
  arrange(desc(Conc)) %>%
  select(ID, Year, Conc) %>%
  head(5)
```

Accept `slice_head(n = 5)` or `head(..., 5)`; `top_n()` is fine but is not in the chapter.

# Part 4 — Reshape

## 12. Wide to long

```{r pivot-longer}
long_data <- all_data %>%
  pivot_longer(cols = c(micro_PC1, micro_PC2, urine_PC1, urine_PC2),
               names_to = "Measure", values_to = "Score")
head(long_data)
dim(long_data)
```

3,200 rows (800 × 4) and 7 columns (`ID`, `Date`, `Conc`, `Year`, `Month`, `Measure`, `Score`).

```{r score-histograms, warning=FALSE}
ggplot(long_data, aes(x = Score)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(~ Measure) +
  theme_classic() +
  labs(x = "Principal component score", y = "Number of samples",
       title = "Distribution of gut microbiome and urine metabolome scores")
```

Grading: four panels, `theme_classic()`, labels, a title. The `warning=FALSE` suppresses the "rows containing non-finite values" message from the NAs; showing the warning is not a fault.

## 13. Long to wide

```{r pivot-wider}
wide_conc <- all_data %>%
  select(ID, Year, Conc) %>%
  pivot_wider(names_from = Year, values_from = Conc) %>%
  mutate(Change = `2022` - `2021`)
head(wide_conc)
sum(wide_conc$Change > 0, na.rm = TRUE)
sum(is.na(wide_conc$Change))
```

**Question 5:** How many IDs increased, how many are `NA`, and why?

**Answer:** 19 IDs increased and 6 are `NA` (the rest, 375, decreased, which is the year effect from Question 4 seen sample by sample). The NAs are the IDs whose `Conc` is missing in one of the two years because that sample is absent from `Dataset1.csv` (they entered `all_data` only through the full join with the other files), so the subtraction has nothing to work with. Students who joined by `ID` only, or who forgot `select()`, get a `pivot_wider()` warning about duplicate rows and list-columns; send them back to Section 7.

# Part 5 — Test the hypothesis visually

## 14. PC1 against PC2

```{r pc-scatter, warning=FALSE}
ggplot(all_data, aes(x = micro_PC1, y = micro_PC2, colour = Conc)) +
  geom_point() +
  facet_wrap(~ Year) +
  theme_classic() +
  labs(x = "Gut microbiome PC1", y = "Gut microbiome PC2",
       colour = "Viral load (CT)",
       title = "Gut microbiome composition and viral load, by year")

ggplot(all_data, aes(x = urine_PC1, y = urine_PC2, colour = Conc)) +
  geom_point() +
  facet_wrap(~ Year) +
  theme_classic() +
  labs(x = "Urine metabolome PC1", y = "Urine metabolome PC2",
       colour = "Viral load (CT)",
       title = "Urine metabolome composition and viral load, by year")
```

Grading: at least one PC1–PC2 scatter with concentration mapped to colour (or size, or a cut into groups), `theme_classic()`, labels, title. Facets by year are asked for but a single panel with `colour = Conc` and `shape = factor(Year)` also answers the question.

**Question 6:** What do the data say about the hypothesis?

**Answer:** Any careful reading of their own figures. What the figures show: there is a colour gradient across both PC spaces. High viral loads sit at high `micro_PC1` / low `micro_PC2` and at low `urine_PC1` / high `urine_PC2` (correlations of about ±0.45 in 2021, ±0.25 in 2022), so patients with similar loads do tend to have similar gut communities and metabolomes, which supports the hypothesis. The pattern is clearest in 2021; in 2022 almost every load is below 3, so on a colour scale shared across the two facets that panel looks nearly uniform. Students who notice this and plot the years separately, or put concentration on a log scale, deserve credit for it. Full credit for a "no support" conclusion only if it is argued from something visible in their figure; the point is reasoning from the plot, not the verdict.

**Question 7:** The other scenario.

**Answer:** The code is the same line for line; only the column names chosen in Section 5, the axis labels and the interpretation change. That is the transferable-skill point: wrangling, joining, reshaping and plotting are the same operations whether the rows are patients or lakes, so time spent learning them pays off in any field.

# DO THIS LAST {-}

When your document is finished and knitted, copy and paste this into the R Console to stop positpal and create your coding log. **Do not include it as a code chunk.**

```
positpal::stop_capture()
positpal::export_capture()
```

`export_capture()` prints the name and location of the `my-capture-YYYYMMDD-HHMMSS.zip` it creates beside this file. **Each group member** uploads their own `.zip` to the **"CC3 — Coding Log (individual)"** dropbox on OnQ; the group's `.Rmd` and knitted `.html` go to your section's **TUT02-CC3** or **TUT03-CC3** folder. Submissions without the log are incomplete. If `export_capture()` fails, zip the `.capture` folder yourself and upload that instead.
