---
title: "BIOL343 Coding Challenge 4"
author: "ANSWER KEY — for TAs, not for OnQ"
date: "2026-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BIOL343 Coding Challenge 4

## Grading notes {-}

Week 5, RSCC Chapter 1 *Populations and Distributions*. Graded 0–5 for completeness. Expected values below are for the full data (966 plants); students who drop the 5 `NA` umbel angles before Section 4 get the same face-diameter numbers, since those columns have no NAs. Section 7 depends on `set.seed(343)` being run in the same chunk as the three `sample()` calls, in that order; any other seed or order gives different sub-samples but the same qualitative answer. The template no longer names functions in most tasks or cites the textbooks (Rob's edits 2026-09-29; the chapter pointers below are for TAs), so accept any code that produces the required result; Section 5 is the exception, where the task says `dplyr`. The chunk options asked for (`message=FALSE` in 3 and 8, `echo=FALSE` + `fig.cap` in 10) are part of the task.

| Section | Expect |
|---|---|
| 4 | SI: n 866, mean 10.54, SD 1.69, CV 0.161 · SC: n 100, mean 5.96, SD 0.89, CV 0.150 |
| 6 | SE 0.0575 (SI), 0.0893 (SC) · CI 10.43–10.66 (SI), 5.79–6.14 (SC) |
| 7 | n 10 / 100 / 500: mean 10.49 / 10.34 / 10.58 · SD 1.91 / 1.74 / 1.68 · CI width 2.36 / 0.68 / 0.29 (full sample 0.23) |
| 8 | z of the SC mean = −2.71 · P(≤ 5.96 from SI) = 0.0034 |
| 9 | Welch t = 43.1, df ≈ 194, p < 2.2e-16, CI of difference 4.37–4.79 mm; hand-calculated t = 43.1 |

## Group Members

```{r group_members}
# Name - Queen's Student Number

# Key drafted 2026-09-28; synced to Rob's template edits 2026-09-29

```

# DO THIS FIRST {-}

Copy and paste this into your R Console to start positpal, but replace `FILENAME.Rmd` with the name you are going to use for the .R or .Rmd file you are going to upload (rename the template first, e.g. `Group_#_CC4.Rmd`, and use that exact name). **Do not include it as a code chunk.**

```
install.packages("remotes")
remotes::install_url("https://quantitative.bio/downloads/positpal.tar.gz", upgrade = "never")
positpal::start("FILENAME.Rmd")
```

Before you run it: create or open an R project folder so that it is your working directory, then save this file in that folder under the name you will upload, and check that autosave is on (*File → Auto Save*). See OnQ and use the positpal chatbot if you are having trouble with positpal.

# Part 1 — Import and inspect

## 1. Import the data

```{r import-data}
floral_data <- read.csv("data/AbroniaFloralData.csv")
```

## 2. Inspect the data

```{r inspect-data}
dim(floral_data)
head(floral_data)
str(floral_data)
summary(floral_data)
```

966 rows × 8 columns. `self_inc`, `site` and `plant_id` are character; the four traits are numeric or integer. The only missing values are 5 `NA` in `umbel_angle_deg` (all SI plants), already coded as `NA`, so nothing needs correcting (the correction-with-comments step was cut 2026-09-29). Grading: the four calls and a sentence noting the NAs.

**Question 1:** What is the statistical population, and what is the sample? One reason the sample might be biased.

**Answer:** The population is all *Abronia umbellata* plants (or all plants of each mating system) across the species' range; the sample is the 966 plants measured at 12 sites. The sites are not a random draw of the coastline (they are the accessible dune sites the researcher chose), plants within a site may not have been picked at random (easy-to-reach or larger plants), only three sites are SC and all three are at the northern edge, so mating system is confounded with latitude, and the flowers were measured in one season. Accept any one of these.

# Part 2 — Distributions

## 3. Plot frequency distributions

```{r face-histogram, message=FALSE}
library(ggplot2)
ggplot(floral_data, aes(x = face_diam_mm)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(~ self_inc, ncol = 1) +
  labs(x = "Floral face diameter (mm)", y = "Number of plants",
       title = "Face diameter of SI and SC Abronia umbellata") +
  theme_classic()
```

```{r flower-histogram, message=FALSE}
ggplot(floral_data, aes(x = flower_number)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ self_inc, ncol = 1) +
  labs(x = "Flowers per inflorescence", y = "Number of plants",
       title = "Flower number of SI and SC Abronia umbellata") +
  theme_classic()
```

Any `binwidth` between about 0.25 and 1 mm is fine for face diameter; the default 30 bins without a `binwidth` misses the instruction. The task asks for "a 'clean' theme" rather than naming one: `theme_classic()`, `theme_bw()`, `theme_minimal()` or a custom theme all qualify. `ncol = 1` (stacked panels) is nicer for comparing the x-axis but not required. The `message=FALSE` is what silences the `stat_bin()` message (the task calls it a warning; same fix); if it appears in the HTML, the option was not set.

**Question 2:** Which distribution is the best model for each variable?

**Answer:** `face_diam_mm` is a continuous measurement, so **Gaussian**; the histograms are roughly bell-shaped within each mating system, which supports the choice. `flower_number` is a count (non-negative integers), so **Poisson** is the natural model; its histogram is a set of integer bars, mildly right-skewed. The task now asks for both the data type *and* the shape, so a full answer mentions each. Binomial fits neither: it is for TRUE/FALSE outcomes. Bonus credit if a student notices that flower number is *less* variable than a Poisson with the same mean would be (variance/mean ≈ 0.65 in both groups), but that is not required.

## 4. Mean, standard deviation and coefficient of variation

```{r metrics}
SI <- floral_data$face_diam_mm[floral_data$self_inc == "SI"]
SC <- floral_data$face_diam_mm[floral_data$self_inc == "SC"]

# sample sizes
length(SI)
length(SC)

# SI
mean(SI)
sd(SI)

# SC
mean(SC)
sd(SC)

# the same two numbers from the Chapter 1 equations (not required by the task)
sum(SI) / length(SI)
sqrt(sum((SI - mean(SI))^2) / (length(SI) - 1))

# coefficient of variation
sd(SI) / mean(SI)
sd(SC) / mean(SC)
```

SI: n = 866, mean 10.54 mm, SD 1.69 mm, CV 0.161 (16.1%). SC: n = 100, mean 5.96 mm, SD 0.89 mm, CV 0.150 (15.0%). `subset(floral_data, self_inc == "SI")$face_diam_mm` or a `dplyr` `filter()` is equally fine. The task no longer asks for the by-equation calculation (removed 2026-09-29), so `mean()`, `sd()` and `sd()/mean()` are enough; the equation lines in the key are there for a TA who wants to check a student who did it that way. CV as a proportion or as a percentage are both fine.

**Question 3:** SDs differ, CVs are almost the same.

**Answer:** The absolute spread of flower sizes is about twice as large in SI plants, but relative to their means the two groups are equally variable (about 15–16% of the mean). Variation in size scales with size: SC flowers are smaller *and* proportionally less variable, which is exactly what the CV is designed to show and why the SD alone can mislead when means differ.

## 5. All four traits by mating system

```{r trait-means, message=FALSE}
library(dplyr)
floral_data %>%
  group_by(self_inc) %>%
  summarise(n_face  = sum(!is.na(face_diam_mm)),   mean_face  = mean(face_diam_mm),
            n_tube  = sum(!is.na(tube_length_mm)), mean_tube  = mean(tube_length_mm),
            n_umbel = sum(!is.na(umbel_angle_deg)), mean_umbel = mean(umbel_angle_deg, na.rm = TRUE),
            n_flow  = sum(!is.na(flower_number)),  mean_flow  = mean(flower_number))
```

| | SC | SI |
|---|---|---|
| Face diameter (mm) | 100 · 5.96 | 866 · 10.54 |
| Tube length (mm) | 100 · 6.21 | 866 · 13.37 |
| Umbel angle (°) | **99** · 88.3 | **862** · 114.6 |
| Flower number | 100 · 13.3 | 866 · 15.7 |

The task requires `dplyr` (Chapter 10), so expect `group_by()` + `summarise()`. The umbel-angle counts are the check: `n()` gives 100/866 for every trait, which is wrong for umbel angle; `sum(!is.na())` gives 99/862. A student who drops the NA rows first (`filter(!is.na(umbel_angle_deg))`) gets 99/862 for *every* trait and slightly different means for the other three (face SI 10.55, tube SI 13.37); accept that with a comment. A wider `pivot_longer()` + `group_by(self_inc, trait)` layout is equally good.

**Question 4:** Direction and largest relative difference.

**Answer:** Yes: all four traits are smaller in SC plants, as the second hypothesis predicts (less investment in attracting pollinators). The largest *relative* differences are tube length (SC ≈ 46% of SI) and face diameter (≈ 57%); umbel angle (≈ 77%) and flower number (≈ 85%) differ less. Accept "face diameter" as the largest if the student reasons from the absolute gap of 4.6 mm; the point is that they compared the traits.

# Part 3 — From the sample to the population

## 6. Standard error and confidence interval

```{r se-ci}
se_SI <- sd(SI) / sqrt(length(SI))
se_SC <- sd(SC) / sqrt(length(SC))
se_SI
se_SC

CI_SI <- c(mean(SI) - 1.96 * se_SI, mean(SI) + 1.96 * se_SI)
CI_SC <- c(mean(SC) - 1.96 * se_SC, mean(SC) + 1.96 * se_SC)
CI_SI
CI_SC
```

SE 0.058 (SI) and 0.089 (SC); CI 10.43–10.66 mm (SI) and 5.79–6.14 mm (SC). Note the SC standard error is larger despite the smaller SD because n is 100, not 866.

**Question 5:** What the CI represents; overlap.

**Answer:** Each interval is the range within which we expect the **population** mean face diameter to lie, with 95% confidence, given this sample (Chapter 1: a range of predicted population means). The intervals are about 4.3 mm apart and do not come close to overlapping, so the two mating systems almost certainly come from populations with different mean flower sizes. (Non-overlapping CIs imply a significant difference; overlapping ones do not necessarily imply no difference, but that subtlety is not required.)

## 7. The central limit theorem

```{r clt}
set.seed(343)
sub10  <- sample(SI, 10)
sub100 <- sample(SI, 100)
sub500 <- sample(SI, 500)

ci_width <- function(x) 2 * 1.96 * sd(x) / sqrt(length(x))

c(mean = mean(sub10),  sd = sd(sub10),  ci_width = ci_width(sub10))
c(mean = mean(sub100), sd = sd(sub100), ci_width = ci_width(sub100))
c(mean = mean(sub500), sd = sd(sub500), ci_width = ci_width(sub500))
c(mean = mean(SI),     sd = sd(SI),     ci_width = ci_width(SI))
```

With `set.seed(343)` and this order: n = 10 → mean 10.49, SD 1.91, CI width 2.36; n = 100 → 10.34, 1.74, 0.68; n = 500 → 10.58, 1.68, 0.29; all 866 → 10.54, 1.69, 0.23. A custom function is not required (three copies of the width formula are fine). Different seeds or order give different numbers with the same pattern; grade the pattern.

**Question 6:** Which number stays put, which shrinks.

**Answer:** The mean and the SD wobble around the full-sample values at every n and settle toward them as n grows (the sample distribution approaches the population distribution — the chapter's statement of the central limit theorem). The CI width shrinks steadily. The hint points at the standard error: $SE = s/\sqrt{n}$, and the CI is $\pm 1.96 \times SE$, so $s$ stays roughly constant while $\sqrt{n}$ grows, and the uncertainty about the population mean falls with the square root of the sample size. Ten times more plants gives a CI about $\sqrt{10} \approx 3.2$ times narrower. Full marks need the SE link, not just "bigger samples are more precise".

# Part 4 — Probability and tests

## 8. Z-scores and probability

```{r z-scores, message=FALSE}
z_SI <- (SI - mean(SI)) / sd(SI)
ggplot() +
  geom_histogram(aes(x = z_SI), binwidth = 0.25) +
  labs(x = "Face diameter (z-score)", y = "Number of plants",
       title = "Standardized face diameter of SI plants") +
  theme_classic()

z_SC <- (mean(SC) - mean(SI)) / sd(SI)
z_SC
pnorm(q = z_SC, mean = 0, sd = 1, lower.tail = TRUE)
# equivalently, without standardizing:
pnorm(q = mean(SC), mean = mean(SI), sd = sd(SI), lower.tail = TRUE)
```

z = −2.71; P = 0.0034. The z histogram must have the same shape as the SI panel in Section 3 with the x-axis re-scaled to SD units, centred on 0. The common bracket error `SI - mean(SI) / sd(SI)` gives a histogram centred near 4, not 0. `lower.tail = TRUE` (the default) is the correct choice here: the question is "that small or smaller".

**Question 7:** One- or two-tailed; interpretation.

**Answer:** One-tailed: the hypothesis predicts a direction (SC flowers *smaller*), so we ask only about the lower tail. About 0.3% of SI plants have a face diameter as small as the *average* SC plant, so the typical SC flower would be an extreme outlier in the SI population; SC flowers are not just small SI flowers. Accept a student who says two-tailed and doubles the probability to 0.007, if they explain why.

## 9. Two-sample *t*-test

```{r t-test}
t.test(x = SI, y = SC)

# by hand, from the two-sample equation
(mean(SI) - mean(SC)) / sqrt(se_SI^2 + se_SC^2)
```

Welch t = 43.15, df = 194.45, p < 2.2e-16, 95% CI of the difference 4.37–4.79 mm; the hand calculation gives 43.15 as well. The task says "a t-test" without naming the function: `t.test(x = SI, y = SC)` and the formula interface `t.test(face_diam_mm ~ self_inc, data = floral_data)` (same result, sign reversed: SC − SI) are both fine. A one-sample or paired test is wrong here. Grading: the call, the four reported numbers, and the matching hand calculation.

**Question 8:** Two assumptions and whether the data meet them.

**Answer:** (1) The populations are Gaussian: plausible for a continuous size trait, and the Section 3 histograms are roughly bell-shaped, though the SC sample is small. (2) The sample is random and unbiased: doubtful, for the reasons in Question 1 (chosen sites, SC confounded with latitude). Also acceptable: independence of observations (plants from the same site are not independent of each other, which a *t*-test ignores) and equal variances (clearly violated, SD 1.69 vs 0.89, which is why R uses the Welch correction and why df is not a whole number).

## 10. Face diameter against tube length

```{r face-vs-tube, echo=FALSE, fig.cap="Figure 1. Floral tube length against floral face diameter for 966 Abronia umbellata plants from 12 coastal sites. Each point is one plant; colour shows whether the plant's site is self-incompatible (SI, n = 866) or self-compatible (SC, n = 100)."}
ggplot(floral_data, aes(x = face_diam_mm, y = tube_length_mm, colour = self_inc)) +
  geom_point(alpha = 0.6) +
  labs(x = "Floral face diameter (mm)", y = "Floral tube length (mm)",
       colour = "Mating system") +
  theme_classic()
```

Check the knitted HTML: no code visible for this chunk, a caption under the figure. Any clean theme is fine (the task no longer names `theme_classic()`). `geom_smooth()` is not asked for (Chapter 8 is not assigned) but is not penalized.

**Question 9:** The two hypotheses.

**Answer:** Hypothesis 1 (SC evolves toward range limits): supported. The three SC sites (CFAA, CGBA, COSA) are the three northernmost, at 40.8–41.4 °N, near the species' northern limit in Oregon; all nine SI sites are further south (34.2–37.0 °N). Hypothesis 2 (SC plants invest less in attracting pollinators): supported. Every trait is smaller in SC plants, the face-diameter difference is far beyond sampling error (non-overlapping CIs, t = 43, p ≈ 0), and the scatterplot shows SC plants as a separate cluster with both face diameter and tube length reduced together, consistent with a general reduction in flower size rather than a change in one trait. A good answer also notes the caveat from Questions 1 and 8: with only three SC sites, all at one end of the range, mating system and latitude cannot be separated, so latitude (or climate) rather than selfing could explain the smaller flowers.

# DO THIS LAST {-}

When your document is finished and knitted, copy and paste this into the R Console to stop positpal and create your coding log. **Do not include it as a code chunk.**

```
positpal::stop_capture()
positpal::export_capture()
```

`export_capture()` prints the name and location of the `my-capture-YYYYMMDD-HHMMSS.zip` it creates beside this file. **Each group member** uploads their own `.zip` to the **"CC4 — Coding Log (individual)"** dropbox on OnQ; the group's `.Rmd` and knitted `.html` go to your section's **TUT02-CC4** or **TUT03-CC4** folder. Submissions without the log are incomplete. If `export_capture()` fails, zip the `.capture` folder yourself and upload that instead.
